RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-07-31 manuscripts/manuscript.Rmd
edits to address #81 and #82

The R language `r citep(citation())` has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `phylo` format provided by the
ape package `r citep(citation("ape"))`.

@@

<!-- actually not yet, but before submission -->

@@

<!-- 
The latest stable development version can be installed from Github using the `devtools` package:

```r
library(devtools)
install_github("RNeXML", "ropensci", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
```
-->

@@

<!-- I still question the wisdom of having multiple names for the same
method. At first this might seem user-friendly, but it will just be confusing
and create a document and source code maintenance nightmare. Why not just pick
read.nexml and be done with it? -->
We begin by reading in an example NeXML file provided with the package.[^2]

<!-- I think it would be useful to provide an overview of what is contained in this -->
<!-- example file here. The reader will know better what to expect (e.g., the -->
<!-- object contains 2 trees.) -->

```{r}
f <- system.file("examples", "trees.xml", package="RNeXML")
nex <- nexml_read(f)
```

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `phylo` object from the 
popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format:

@@

<!-- I am not sure it really matters but `phy` is actually a `multiPhylo` -->
<!-- object -->

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

@@

Like NEXUS files, NeXML documents can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological) data 
used to infer the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to extract character data
work much like those for the phylogenetic data:

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```
<!-- I think it's better to be explicit rather than referring to the chunk above -->

`get_characters` returns a `data.frame` with columns as characters and rows as
taxa. Continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

<!-- in caper there is this method comparative.data to join data
and trees on taxon labels. it would be nice if this merged data
structure could also be passed to nexml_write (or should that be
write.nexml?) -->
The first argument to `nexml_write` can be a `phylo` phylogeny,
`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
can be coerced into the `nexml` class.  As we have seen, NeXML files
can contain multiple phylogenetic trees, as well as character matrices.
We can add both trees  and character data using `nexml_write`,

@@

<!-- how about add_taxa? -->
NeXML files can also be built up bit by bit using the `add_characters`,
`add_trees`, and `add_meta`.  Each of these functions can add data to an
existing `nexml` object or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

<!-- in the Perl library I also embed the value of the environment variable
$USER as a document-level annotation -->
`RNeXML` automatically adds basic metadata by default such as a time-stamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions

@@

This function uses the R package `taxize` `r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation <!-- in a way a computer
can understand -- this is redundant with the beginning of the sentence -->.  For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML file as:

@@

Using this convention we can generate NeXML files containing simmap
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a `nexml` object containing
a simmap reconstruction into R:

@@

The `get_trees` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the simmap reconstruction data
and returns includes this in a `maps` element of the `phylo` object,
for use with other `phytools` functions.

@@

We can then use various functions from `phytools` designed for simmap
objects `r citep(citation("phytools"))`, such as the plotting function:

@@

Though other NeXML parsers (for instance, for Perl or Python) have
not been written explicitly to express simmap data, those parsers will
nonetheless be able to successfully parse this file and expose the simmap
data to the user.

>>>>

The R language `r citep(citation())` has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package `r citep(citation("ape"))`.

@@

@@

@@

<!-- I still question the wisdom of having multiple names for the same
method. At first this might seem user-friendly, but it will just be confusing
and create a document and source code maintenance nightmare. Why not just pick
read.nexml and be done with it? -->
<!-- For whatever reason, R has native support for aliasing function names;
     so I don't see this creating a maintenance/documentation issue.  If I 
     had to pick, I'd go with read_nexml though; a lot of the R developer 
     community strongly opposes the `.` in function names, since it falsely
     mimics the S3 class structure.  - CB -->

We begin by loading the `RNeXML` package and reading in an example
NeXML file provided with the package, containing two small phylogenetic
trees.[^2]

```{r}
library("RNeXML")
f <- system.file("examples", "trees.xml", package="RNeXML")
nex <- nexml_read(f)
```

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `ape::phylo` object from the 
popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

@@

Like NEXUS files, NeXML documents can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological) data 
used to infer the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to extract character data
work much like those for the phylogenetic data. Here we read in an example
NeXML file that contains both a phylogeny and character data. 
`get_characters` returns a `data.frame` with columns as characters and rows as
taxa:

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```

As this example shows, continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

<!-- in caper there is this method comparative.data to join data
and trees on taxon labels. it would be nice if this merged data
structure could also be passed to nexml_write (or should that be
write.nexml?) -->
<!-- Great observation.  I think we should add a coercion method
from the caper structure to the nexml S4 structure.  Then nexml_write
can automatically handle this. added as issue now  -->

The first argument to `nexml_write` can be a `ape::phylo` phylogeny,
`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
can be coerced into the `nexml` class.  As we have seen, NeXML files
can contain multiple phylogenetic trees, as well as character matrices.
We can add both trees  and character data using `nexml_write`,

@@

<!-- how about add_taxa? -->
<!-- A taxa block is added when trees or characters are added.  Yeah,
     I guess it would make sense to be able to add taxa directly,
     (e.g. a NeXML file containing only RDFa metadata about taxa, 
     independent of characters or trees), but I worry about confusing
     people into thinking they must add taxa blocks separately, and
     about when we should merge or not merge taxa blocks of matching
     species. See https://github.com/ropensci/RNeXML/issues/43 
     Added new issue for this, https://github.com/ropensci/RNeXML/issues/85 -->

NeXML files can also be built up bit by bit using the `add_characters`,
`add_trees`, and `add_meta`.  Each of these functions can add data to an
existing `nexml` object or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

<!-- in the Perl library I also embed the value of the environment variable
$USER as a document-level annotation -->
<!-- Good idea-->

`RNeXML` automatically adds basic metadata by default such as a time-stamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions.

The `taxize_nexml` function uses the R package `taxize` 
`r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

@@

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation. For instance, the title we gave to the `bird.orders`
phylogeny appears in the NeXML file as:

@@

Using this convention we can generate NeXML files containing `simmap`
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a `nexml` object containing
a `simmap` reconstruction into R:

@@

The `get_trees` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the `simmap` reconstruction data
and returns includes this in a `maps` element of the `ape::phylo` object,
for use with other `phytools` functions.

@@

We can then use various functions from `phytools` designed for `simmap`
objects `r citep(citation("phytools"))`, such as the plotting function:

@@

Though other NeXML parsers (for instance, for Perl or Python) have
not been written explicitly to express `simmap` data, those parsers will
nonetheless be able to successfully parse this file and expose the `simmap`
data to the user.
